---
title: "Blog: Week 2 - Economy"
output:
  md_document:
    variant: markdown_github
date: '2022-09-18'
---

```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(tidyverse)
library(ggplot2)
library(knitr)
```
\tiny

Using economic data is a key element in many models used to predict presidential or midterm elections. People associate politics with their personal finances, and the media constantly reports about inflation, unemployment, and GDP figures. While the President does not have full control over the economy, many Americans end up using this information as a way to criticize or applaud the current administration.But, as we will see, the economy may not be as useful as originally thought in predicting Republican or Democratic vote share in the midterm elections. 

```{r, echo=FALSE,message=FALSE}
#read in the data
cpi_monthly = read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week2/CPI_monthly.csv")
GDP_quarterly = read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week2/GDP_quarterly.csv")

unemployment_national = read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week2/unemployment_national_quarterly_final.csv")

unemployment_state_monthly = read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week2/unemployment_state_monthly.csv") %>% 
  filter((Year %%2 ==0) & (Month==10)) %>%
  select("State and area","Year","Unemployed_prct") %>%
  filter(Year %in% c(2018,2014,2010,2006,2002,1998,1994,1990,1986,1982,1978,1974,1970,1966,1962))

cpi_df = cpi_monthly %>%
  mutate(Year = as.numeric(str_split(cpi_monthly$DATE,"-",simplify=T)[,1]),
         Month = as.numeric(str_split(cpi_monthly$DATE,"-",simplify=T)[,2])) %>% 
  group_by(Month) %>% arrange(DATE) %>%
  mutate(Inflation_yoy=100*(CPIAUCSL - lag(CPIAUCSL,1))/lag(CPIAUCSL,1)) %>%
  ungroup() %>% arrange(DATE) %>%
  filter((Year %% 2 == 0) & (Month ==10))

gdp_df = GDP_quarterly %>%
  filter((year %% 2 == 0) & (quarter_yr ==4)) %>% rename(Year=year)

unemployment_national_df = unemployment_national %>%
  filter((year %% 2 == 0) & (quarter_yr == 4))%>% rename(Year=year) %>%
  select(Year,UNRATE)

house_popvote = read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week2/house_popvote_seats.csv") %>% rename(Year=year)

df = left_join(left_join(left_join(house_popvote,cpi_df,by=c("Year")), gdp_df, by="Year"),unemployment_national_df,by="Year")%>% 
  select(Year,
         R_seats,
         D_seats,
         winner_party,
         R_majorvote_pct,
         D_majorvote_pct,
         president_party,
         H_incumbent_party_winner,
         H_incumbent_party_majorvote_pct,
         H_incumbent_party,
         Inflation_yoy,
         GDP_growth_pct,
         UNRATE) %>% na.omit() %>%
  filter(Year %in% c(2018,2014,2010,2006,2002,1998,1994,1990,1986,1982,1978,1974,1970,1966,1962)) %>%
  mutate(president_party_house_vote = case_when(
           president_party == "D" ~ D_majorvote_pct,
           president_party == "R" ~ R_majorvote_pct
  ))
```

```{r,echo=FALSE,message=FALSE}
#read in the data
cpi2022 = read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week2/CPI_monthly.csv") %>%
  mutate(Year = as.numeric(str_split(cpi_monthly$DATE,"-",simplify=T)[,1]),
         Month = as.numeric(str_split(cpi_monthly$DATE,"-",simplify=T)[,2])) %>% 
  group_by(Month) %>% arrange(DATE) %>%
  mutate(Inflation_yoy=100*(CPIAUCSL - lag(CPIAUCSL,1))/lag(CPIAUCSL,1)) %>%
  ungroup() %>% arrange(DATE) %>%
  filter((Year == 2022)& (Month==6))

gdp2022= read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week2/GDP_quarterly.csv") %>%
  filter((year == 2022) &(quarter_yr == 2)) %>% rename(Year=year)

unemployment2022 = read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week2/unemployment_national_quarterly_final.csv") %>%
  filter((year == 2022) &(quarter_yr == 2))%>% rename(Year=year) %>% select(Year,UNRATE)

economy.2022 = left_join(left_join(cpi2022, gdp2022, by="Year"),unemployment2022,by="Year") %>% 
  select(Year,
         Inflation_yoy,
         GDP_growth_pct,
         UNRATE) 
```

The first logical step in creating a predictive model for the 2022 Midterm Elections is to visualize how economic indicators have related to election outcomes in the past. One key decision in creating these variables is determining which inflation, unemployment, and GDP growth metrics to use. For example, a forecast may decide to use GDP growth between the first fiscal quarter of a President's term through to the 7th fiscal quarter (which is when the midterms take place) while others may use GDP growth statistics from only the election year  Borrowing from Achen and Bartels' "Retrospective Voting" concept, voters make choices based on how the economy has performed. However, many people forget how the economy has done throughout the first two years of a President's term, instead only focusing on how the economy has done in the months leading up to the election. Therefore, it makes sense to take the most recent economic information before the election and using the measurement that the media commonly reports (because this is how most voters consume this information). Combining these two ideas, the plots depict year over year [inflation](https://www.bls.gov/news.release/pdf/cpi.pdf) (media outlets report the year-over-year number) and [unemployment](https://www.bls.gov/news.release/pdf/empsit.pdf) for the month of October before a midterm. Finally, GDP growth between the second and third fiscal quarter of the election year gives the most recent performance. As seen below, none of these metrics seem to have a clear linear relationship with how the the sitting President's party performs in the midterms. Besides 1962, 1966, 1978, and 2002, every party with control over the white house lost a majority of the vote share regardless of economic factors. 

```{r,echo=FALSE}
df %>%
  ggplot(aes(x=Inflation_yoy, y=president_party_house_vote,
             label=Year)) + 
    geom_text() +
    geom_hline(yintercept=50, lty=2) + # median
    xlab("YoY Inflation in October of the Election Year") +
    ylab("Congressional Vote Share Nationally") +
  ggtitle("Inflation vs Current President party House Vote Share")+
    theme_bw() +
    theme(
      axis.text = element_text(size = 10))

# ggsave("InflationVsHouseVoteShare.png", height = 6, width = 12)


df %>%
  ggplot(aes(x=UNRATE, y=president_party_house_vote,
             label=Year)) + 
    geom_text() +
    geom_hline(yintercept=50, lty=2) + # median
    xlab("Unemployment in October of the Election Year") +
    ylab("Congressional Vote Share Nationally") +
  ggtitle("Unemployment vs Current President party House Vote Share")+
    theme_bw() +
    theme(
      axis.text = element_text(size = 10))
# ggsave("UnemploymentVsHouseVoteShare.png", height = 6, width = 12)

df %>%
  ggplot(aes(x=GDP_growth_pct, y=president_party_house_vote,
             label=Year)) + 
    geom_text() +
    geom_hline(yintercept=50, lty=2) + # median
    xlab("Q2-Q3 GDP Growth of the Election Year") +
    ylab("Congressional Vote Share Nationally") +
  ggtitle("GDP Growth vs Current President party House Vote Share")+
    theme_bw() +
    theme(
      axis.text = element_text(size = 10))

# ggsave("GDPvsHouseVoteShare.png", height = 6, width = 12)

```

\tiny

**Creating Linear Models**

With this in mind, I still wanted to create simple linear models to predict Republican vote share. I decided to exclude Presidential elections for two main reasons. First, turnout in Presidential elections is normally considerably higher than in [midterm elections](https://fairvote.org/resources/voter-turnout/). Second, Presidential candidates have much larger national profiles compared to the average congressional representative, so there is much more media coverage. Ultimately, I think the dynamics of Presidential elections are too different, so it is safer to exclude them all together. 

\tiny

At first, I tried creating separate linear models for inflation, GDP growth rate, and unemployment as sole independent variables to predict each party's vote share (party votes plus third party candidates who lean in the same direction). The results show that only inflation is a significant predictor (its p-value is less than $0.05$). When predicting Republican vote share, the GOP are predicted to lose $0.75$% of the popular vote share for every $1$% increase in inflation. This finding is the opposite for Democrats, which gain a predicted $0.75$% in the popular vote share for every $1$% increase in inflation. These results are not compelling, and while the plot below appears to suggest there is a positive linear relationship between inflation and Democratic vote share, many of these elections require context (the most obvious example is in 1974 when President Ford pardoned Richard Nixon after the Watergate Scandal two months before the midterms). Also, the inflation model predicts Democrats win $56$% of the vote share which is extremely optimistic and unlikely due to President Biden's poor approval ratings. More generally, using past midterms when there are only $15$ observations in the data set means that small changes in past election outcomes could lead to very different results in the models. 
```{r,echo=FALSE,message=FALSE}
df %>%
  ggplot(aes(x=Inflation_yoy, y=D_majorvote_pct,
             label=Year)) + 
    geom_text(size = 5) +
    geom_smooth(method="lm", formula = y ~ x) +
    geom_hline(yintercept=50, lty=2) +
    geom_vline(xintercept=0.0, lty=2) + # median
    xlab("Inflation") +
    ylab("Democrat Popular Vote Share") +
    theme_bw() +
    theme(axis.text = element_text(size = 10),
          axis.title = element_text(size = 10),
          plot.title = element_text(size = 10))

# ggsave("InflationvsDemocrats.png", height = 6, width = 12)
r_inflation_model = lm(R_majorvote_pct ~ Inflation_yoy,data = df)
d_inflation_model = lm(D_majorvote_pct ~ Inflation_yoy,data = df)

r_gdp_model = lm(R_majorvote_pct ~ GDP_growth_pct,data = df)
d_gdp_model = lm(D_majorvote_pct ~ GDP_growth_pct,data = df)


r_unemployment_model = lm(R_majorvote_pct ~ UNRATE,data = df)
d_unemployment_model = lm(D_majorvote_pct ~ UNRATE,data = df)

# at most 4 decimal places
kable(cbind(Metric.Republican_Voteshare = c("Inflation","GDP Growth","Unemployment"),
  Coefficient = c(
    round(summary(r_inflation_model)$coef[2],digits=3),
    round(summary(r_gdp_model)$coef[2],digits=3),
    round(summary(r_unemployment_model)$coef[2], digits=3)), 
  P.Value = c(
    round(summary(r_inflation_model)$coef[8],digits=3),
    round(summary(r_gdp_model)$coef[8],digits=3),
    round(summary(r_unemployment_model)$coef[8],digits=3)),
  Predictions = c(
    round(predict(r_inflation_model, economy.2022), digits=3),
    round(predict(r_gdp_model,economy.2022), digits=3),
    round(predict(r_unemployment_model,economy.2022), digits=3)
  )))



kable(cbind(Metric.Democrat_Voteshare = c("Inflation","GDP Growth","Unemployment"),
  Coefficient = c(
    round(summary(d_inflation_model)$coef[2],digits=3),
    round(summary(d_gdp_model)$coef[2],digits=3),
    round(summary(d_unemployment_model)$coef[2], digits=3)), 
  P.Value = c(
    round(summary(d_inflation_model)$coef[8],digits=3),
    round(summary(d_gdp_model)$coef[8],digits=3),
    round(summary(d_unemployment_model)$coef[8],digits=3)),
  Predictions = c(
    round(predict(d_inflation_model, economy.2022), digits=3),
    round(predict(d_gdp_model,economy.2022), digits=3),
    round(predict(d_unemployment_model,economy.2022), digits=3)
  )))


```


\

Next, I asked a different question. Instead of predicting how many votes Republicans or Democrats will win, the better question may be less concerned with specific party. Put more generally, how will the current President's party perform in the midterms based on economic factors? This makes more sense because voters punish the current President for bad economic performance. When looking at it this way, none of the predictors are significant, but the coefficients remain interpretable. In the table below, rises in inflation and unemployment hurt the sitting administration's party while GDP growth has a positive relationship with the party who controls the White House. Using the most recent economic data for 2022, the models return a wide range of predictions for how well President Biden's Party (Democrats) will perform this fall. The inflation model thinks voters will vote overwhelmingly for Republicans (which is far more reasonable than the previous model that heavily favored Democrats because of high inflation), while strong GDP growth indicates the Democrats could win the popular vote. Compared to the previous models, changing the dependent variable reveals that usually, the party of the current President struggles in the midterms.  While these models are far from perfect, they serve as a baseline that we can improve upon immensely. As we move closer to the election, it will be important to account for fear of a recession due to the Federal Reserve's tightening of interest rates.  
```{r,echo=FALSE}

inflation_lm = lm(president_party_house_vote ~ Inflation_yoy,data = df)
gdp_lm = lm(president_party_house_vote ~ GDP_growth_pct,data = df)
unemployment_lm = lm(president_party_house_vote ~ UNRATE,data = df)

models = c("Inflation","GDP Growth","Unemployment")
predictions.2022 = c(round(predict(inflation_lm,economy.2022),digits=3),round(predict(gdp_lm,economy.2022),digits=3),round(predict(unemployment_lm,economy.2022),digits=3))

coefficients = c(round(summary(lm(president_party_house_vote ~ Inflation_yoy,data = df))$coef[2],digits=3),
          round(summary(lm(president_party_house_vote ~ GDP_growth_pct,data = df))$coef[2],digits=3),
          round(summary(lm(president_party_house_vote ~ UNRATE,data = df))$coef[2], digits=3))

p.values = c(
    round(summary(lm(president_party_house_vote ~ Inflation_yoy,data = df))$coef[8],digits=3),
    round(summary(lm(president_party_house_vote ~ GDP_growth_pct,data = df))$coef[8],digits=3),
    round(summary(lm(president_party_house_vote ~ UNRATE,data = df))$coef[8],digits=3))

kable(cbind(models,coefficients,p.values,predictions.2022))
```



