---
title: "swing_states"
output: 
  md_document:
    variant: markdown_github
date: '2022-09-11'
---


***Blog - Week 1***
```{r, echo=FALSE, warning=FALSE,message=FALSE}
#load libraries and data
library(tidyverse)
library(usmap)
library(ggplot2)

# Read in datasets
pvstate_df <- read_csv("../data/week1/house party vote share by district 1948-2020.csv")
```

One common way to predict upcoming midterm congressional elections is by looking at the vote share for both major parties in each state over time. Below are maps of the United States highlighted with different colors depending on the percentage of voters that voted for Republicans, calculated by taking the total number of votes to Republican congressional candidates in the state divided by the total number of votes cast. Red indicates that Republicans carried the state, while blue shades mean it went to the Democrats. White shades indicate that the vote was very close.\

```{r,echo=FALSE}
margin = pvstate_df %>%
  filter(raceYear>=1980) %>%
    select(raceYear, State, RepVotes, DemVotes) %>%
  # summarize party vote share by state
  group_by(State,raceYear) %>%
# mutate Rep vote margin by state %>%
    mutate(R_votemargin_st = (sum(RepVotes))/
                            sum(RepVotes + DemVotes) - (sum(DemVotes))/sum(RepVotes + DemVotes)) %>%
  select(raceYear,State, R_votemargin_st) %>%
    rename(state = State)%>% distinct() %>% filter(raceYear == 2018 | 
                                                     raceYear == 2014 | 
                                                     raceYear == 2010 |
                                                     raceYear == 2006 | 
                                                     raceYear == 2002 | 
                                                     raceYear == 1998 |
                                                     raceYear == 1994 | 
                                                     raceYear == 1990 | 
                                                     raceYear == 1986 | 
                                                     raceYear == 1982)
```



**Vote margin by state 1982-2018 Midterms**
```{r, echo=FALSE}
plot_usmap(data = margin, regions = "state", values = "R_votemargin_st") +
  facet_wrap(facets = raceYear~.) +
  scale_fill_gradient2(low = "blue",mid = "white",high = "red",name = "Vote Share Margin") + 
  labs(title = "Vote Share for Two Major Parties in Midterm Elections",subtitle = "Red corresponds to a larger vote share for Republicans and blue corresponds to a larger vote share for Democrats") + 
  theme(legend.position = "right",
        panel.grid = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 15))
```


We can see from the plots that most states rarely change their party preference. For example, California has voted heavily for Democrats to represent them in Congress since the 1990s while Texas has voted for Republicans in the same time frame. More generally, southern states in the "Sun" or "Bible" Belt have consistently voted for Republicans by large margins while coastal regions with denser urban areas like New England and the West Coast have supported Democrats. While much of midterm election results can be attributed to district level factors and how the congressional maps were drawn, looking at the few states that regularly flip party preferences can answer important questions for national politics. Namely, are there states that used to lean a certain direction that are now becoming more competitive? If certain states are beginning to shift their partisan slant, what do we need to understand about these communities who may end up who holds the power in Congress or in the White House? Lastly, which candidates should the RNC and DNC allocate resources and attention to?\


```{r, echo=FALSE}
swing_state_df <- pvstate_df %>%
    filter(raceYear >= 1976) %>%
    select(raceYear,State,RepVotes, DemVotes) %>%
  group_by(State,raceYear) %>%
  mutate(RepVotesTotal = sum(RepVotes),DemVotesTotal = sum(DemVotes)) %>%
  select(raceYear,State,RepVotesTotal,DemVotesTotal) %>% distinct() %>%
  mutate(partisanship = RepVotesTotal/(RepVotesTotal + DemVotesTotal)) %>%
  group_by(State) %>%
  mutate(swing = case_when(
    raceYear == 1986 ~ partisanship - partisanship[raceYear==1982],
    raceYear == 1990 ~ partisanship - partisanship[raceYear==1986],
    raceYear == 1994 ~ partisanship - partisanship[raceYear==1990],
    raceYear == 1998 ~ partisanship - partisanship[raceYear==1994],
    raceYear == 2002 ~ partisanship - partisanship[raceYear==1998],
    raceYear == 2006 ~ partisanship - partisanship[raceYear==2002],
    raceYear == 2010 ~ partisanship - partisanship[raceYear==2006],
    raceYear == 2014 ~ partisanship - partisanship[raceYear==2010],
    raceYear == 2018 ~ partisanship - partisanship[raceYear==2014]))%>%na.omit() 

colnames(swing_state_df) = c("raceYear","state","RepVotesTotal","DemVotesTotal","partisanship","swing")
```



**Swing Analysis, Midterm Elections 1982-2018**
```{r,echo=FALSE}
plot_usmap(data = swing_state_df, regions = "state", values = "swing") +
  facet_wrap(facets = raceYear~.) +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       name = "Swing Margin") + 
  labs(
    title = "Swing States for U.S. Presidential Elections over Time",
    subtitle = "Positive Numbers corresponds to a larger vote share for Republicans compared to the last election") + 
  theme(legend.position = "right",
        panel.grid = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 15))

# ggsave("swing.png", height = 6, width = 12)
```

Before we answer these questions, it is important to figure out which states are becoming more competitive. From the maps below, the different shades of blue and red represent a “swing” factor. This is calculated by taking the vote share won by the Republicans and subtracting their vote share in the midterm election four years ago. Negative numbers, corresponding to blue, mean that Republicans won a smaller share of the votes compared to four years ago. States that are a deeper shade of red means Republicans won a larger share of the vote than four years ago. Looking at the maps, we can see that Pennsylvania, Michigan, Ohio, and Florida have flipped between red and blue in the past few midterm elections (this also holds true of Presidential elections). There is clear evidence for this as well because politicians at the top of the ballot held most of their campaign rallies in the “rust” belt in the lead up to the 2020 election. According to the [Washington Post](https://www.washingtonpost.com/elections/2020/11/02/campaign-rallies-covid/), Pennsylvania and Florida were the most common destination for both Joe Biden and Donald trump. 

More interesting still, some historically conservative states are experiencing demographic shifts in that could change the political landscape. In particular, [Georgia](https://www.nbcnews.com/politics/meet-the-press/meet-2022-s-most-important-swing-state-georgia-n1287686), [Arizona](https://www.nysun.com/article/why-arizona-is-turning-blue), and [Texas](https://www.theguardian.com/us-news/commentisfree/2022/mar/08/can-texas-become-purple-that-may-depend-on-hispanic-voters) all have rapidly growing cities and surrounding suburbs (which typically vote for Democrats). In addition, immigrants escaping violence from South America naturalizing in Arizona and Texas  could open the door for Democratic gains in the House or Senate. 

