---
title: "swing_states"
output: 
  md_document:
    variant: markdown_github
date: '2022-09-11'
---


***Blog - Week 1***
```{r, echo=FALSE, warning=FALSE,message=FALSE}
#load libraries and data
library(tidyverse)
library(usmap)
library(ggplot2)

# Read in datasets
pvstate_df <- read_csv("../data/week1/house party vote share by district 1948-2020.csv")
```

One common way to predict upcoming midterm congressional elections is to look at the vote share for each major party by state over time. Below are maps of the United States. The highlighted colors represent the proportion of Republican votes calculated by taking the total number of votes for Republican congressional candidates in the state divided by the total number of votes cast. Then, we subtract this number by the proportion of Democratic votes, thus getting a value between $-1$ and $1$. Red indicates that Republicans carried the state, while blue shades reflect a Democratic victory. The darker the shade means that party won by a larger margin. White and pale shades indicate that the vote was very close.\

```{r,echo=FALSE}
margin = pvstate_df %>%
  filter(raceYear>=1980) %>%
    select(raceYear, State, RepVotes, DemVotes) %>%
  # summarize party vote share by state and year
  group_by(State,raceYear) %>%
# get republican vote share
    mutate(R_votemargin_st = (sum(RepVotes))/
                            sum(RepVotes + DemVotes) - (sum(DemVotes))/sum(RepVotes + DemVotes)) %>%
  select(raceYear,State, R_votemargin_st) %>%
  #filter only for midterm years
    rename(state = State)%>% distinct() %>% filter(raceYear == 2018 | 
                                                     raceYear == 2014 | 
                                                     raceYear == 2010 |
                                                     raceYear == 2006 | 
                                                     raceYear == 2002 | 
                                                     raceYear == 1998 |
                                                     raceYear == 1994 | 
                                                     raceYear == 1990 | 
                                                     raceYear == 1986 | 
                                                     raceYear == 1982)
```


\
**Vote margin by state 1982-2018 Midterms**\
```{r, echo=FALSE,fig.width=12, fig.height=6}
#plotting Republican Win Share for Each Midterms
plot_usmap(data = margin, regions = "state", values = "R_votemargin_st") +
  facet_wrap(facets = raceYear~.) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    name = "Vote Share Margin") + 
  labs(
    title = "Vote Share for Two Major Parties in Midterm Elections",
    subtitle = "Red corresponds to a larger vote share for Republicans and blue corresponds to a larger vote share for Democrats") + 
  theme(legend.position = "right",
        panel.grid = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 15))

# ggsave("margin.png", height = 6, width = 12)
```


The maps over time reflect that, in most states, party preference rarely changes. For example, California has voted heavily for Democrats to represent them in Congress since the 1990s while Texas has voted for Republicans in the same time frame. More generally, southern states in the “Sun” or “Bible” Belt have consistently voted for Republicans by large margins while coastal regions with denser urban areas such as New England and the West Coast have supported Democrats. While much of midterm election results can be attributed to district level factors and how the congressional maps were drawn, an examination of the few states that regularly change party preferences can answer these important questions for national politics: (1) are there states that used to lean a certain direction that are now becoming more competitive? (2) If certain states are beginning to shift their partisan slant, what do we need to understand about these communities to indicate who may end up holding power in Congress or in the White House? (3) Lastly, to which candidates should the RNC and DNC allocate resources and attention?\


```{r, echo=FALSE}
swing_state_df <- pvstate_df %>%
  #filter for years over 1976
    filter(raceYear >= 1976) %>%
    select(raceYear,State,RepVotes, DemVotes) %>%
  #groupby state and race year
  group_by(State,raceYear) %>%
  mutate(RepVotesTotal = sum(RepVotes),DemVotesTotal = sum(DemVotes)) %>%
  select(raceYear,State,RepVotesTotal,DemVotesTotal) %>% distinct() %>%
  #create "partisanship" variable that represents Republican vote share
  mutate(partisanship = RepVotesTotal/(RepVotesTotal + DemVotesTotal)) %>%
  group_by(State) %>%
  #create swing variable based off of partisanship and lagged 4 year prior partisanship by state
  mutate(swing = case_when(
    raceYear == 1986 ~ partisanship - partisanship[raceYear==1982],
    raceYear == 1990 ~ partisanship - partisanship[raceYear==1986],
    raceYear == 1994 ~ partisanship - partisanship[raceYear==1990],
    raceYear == 1998 ~ partisanship - partisanship[raceYear==1994],
    raceYear == 2002 ~ partisanship - partisanship[raceYear==1998],
    raceYear == 2006 ~ partisanship - partisanship[raceYear==2002],
    raceYear == 2010 ~ partisanship - partisanship[raceYear==2006],
    raceYear == 2014 ~ partisanship - partisanship[raceYear==2010],
    raceYear == 2018 ~ partisanship - partisanship[raceYear==2014]))%>%na.omit() 

colnames(swing_state_df) = c("raceYear","state","RepVotesTotal","DemVotesTotal","partisanship","swing")
```



**Swing Analysis, Midterm Elections 1982-2018**
```{r,echo=FALSE,,fig.width=12, fig.height=6}
#plot swing metric by state
plot_usmap(data = swing_state_df, regions = "state", values = "swing") +
  facet_wrap(facets = raceYear~.) +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       name = "Swing Margin") + 
  labs(
    title = "Swing States for U.S. Presidential Elections over Time",
    subtitle = "Positive Numbers corresponds to a larger vote share for Republicans compared to the last election") + 
  theme(legend.position = "right",
        panel.grid = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 15))

# ggsave("swing.png", height = 6, width = 12)
```

Before we answer these questions, it is important to figure out which states are becoming more competitive. From the maps below, the different shades of blue and red represent a “swing” factor. This is calculated by taking the vote share won by the Republicans and subtracting their vote share in the midterm election four years ago. Negative numbers, corresponding to blue, mean that Republicans won a smaller share of the votes compared to four years ago. For example, in 2018, Texas did not vote overall for Democratic candidates, but the percent of the vote that went to Democrats was higher than 2014. States that are a deeper shade of red means the opposite - Republicans won a larger share of the vote than they did four years prior. Looking at the maps, we can see that Pennsylvania, Michigan, Ohio, and Florida have flipped between red and blue in the past few midterm elections (this also holds true of Presidential elections). There is clear evidence for this as well because politicians at the top of the ballot held most of their campaign rallies in the “rust” belt in the lead up to the 2020 election. According to the [Washington Post](https://www.washingtonpost.com/elections/2020/11/02/campaign-rallies-covid/), Pennsylvania and Florida were the most common destination for both Joe Biden and Donald trump. 

More interesting still, some historically conservative states are experiencing demographic shifts that could change the political landscape. In particular, [Georgia](https://www.nbcnews.com/politics/meet-the-press/meet-2022-s-most-important-swing-state-georgia-n1287686), [Arizona](https://www.nysun.com/article/why-arizona-is-turning-blue), and [Texas](https://www.theguardian.com/us-news/commentisfree/2022/mar/08/can-texas-become-purple-that-may-depend-on-hispanic-voters) all have rapidly growing cities and surrounding suburbs (which typically vote for Democrats). In addition, immigrants escaping violence from South America living in Arizona and Texas could open the door for Democratic gains in the House and Senate. These demographic changes could have huge consequences for the future of both political parties and for Presidential Elections. Arizona, Georgia, and especially Texas have large electoral college counts, so as these states become more competitive, national organizations and candidates will spend considerably more time, money, and organizational energy attracting “swingable” voters.

