---
title: "week6"
output:
  md_document:
    variant: markdown_github
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/posts"
    )
  })
date: '2022-10-16'
---
**The Ground Game - Week 6**
```{r, echo=FALSE,warning=FALSE,echo=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
library(ggplot2)
library(readxl)
library(ggplot2)
library(blogdown)
library(readr)
library(usmap)
library(rmapshaper)
library(sf)
library(knitr)
```
The “Ground Game”, Political Campaigns’ strategy to mobilize and persuade voters is critically important when trying to win elections. While persuading voters is becoming increasingly difficult as the country becomes more partisan, both political parties are realizing that success is predicated on turnout. There have been many large scale “get-out-the-vote” campaigns in recent years. For example, in 2018 during the Gubnatorial race, Stacey Abrams’s Campaign helped register hundreds of thousands of [voters](https://www.independent.co.uk/voices/georgia-election-stacey-abrams-biden-b1675670.html)
```{r, echo=FALSE,warning=FALSE,echo=FALSE, message=FALSE}

#turnout 
cvap = read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week6/cvap_district_2012-2020_clean.csv") %>% rename(st_cd_fips = geoid) %>% select(st_cd_fips, state, year, cvap)


house_polls = read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week6/house_polls_long.csv")

#had to make my own version of house polls long because there were issues with some districts having vectors of results
polls_df <- read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week4/dist_polls_2018-2022.csv")

polls_df <- polls_df %>%
  select(pollster, sponsors, display_name, fte_grade,
         start_date, end_date, sample_size, population, cycle, election_date,
         party, candidate_name, pct, state, st_fips, cd_fips, st_cd_fips)

# new party candidate name variable
polls_df <- polls_df %>%
  mutate(dem_cand = case_when(party == 'DEM' ~ candidate_name),
         rep_cand = case_when(party == 'REP' ~ candidate_name)) %>%
  filter(party == "DEM" | party == "REP")

# pivot wide
polls_df_wide <- polls_df %>%
  pivot_wider(names_from = party, values_from = pct, values_fn = "mean")

# make long
polls_df_wide <- polls_df_wide %>% 
  fill(dem_cand, rep_cand, .direction = "up") %>%
  fill(DEM, REP, .direction = "up") 

# delete duplicates
ind <- seq(1, nrow(polls_df_wide), by=2)
polls_df_wide <- polls_df_wide[-ind, ]

polls_df_wide = polls_df_wide %>%
  select(cycle,state,st_fips,cd_fips, st_cd_fips, dem_cand, rep_cand, DEM, REP) %>% 
  filter(state != "Alaska")  %>%
  rename(year = "cycle")

```

```{r,  cache=TRUE, eval=TRUE, echo = FALSE, message=FALSE, warning=FALSE}
## Get relevant data
# CVAP

cvap_district <- read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week6/cvap_district_2012-2020_clean.csv")

# rename geoid
cvap_district <- cvap_district %>%
  rename(st_cd_fips = geoid) 

# district-level polls

polls_df <- read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week6/house_polls_long.csv")
polls_df <- polls_df %>%
  filter(year == '2018' | year == '2020' | year == "2022")

# join
cvap_district <- cvap_district %>%
  # filter to relevant years 
  filter(year == '2018' | year == '2020' | year == "2022")

polls_cvap_df <- merge(polls_df, cvap_district, by = c('st_cd_fips', 'year'))

# merge with district-level voteshares
dist_pv_df <- read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week6/incumb_dist_1948-2020.csv")
dist_pv_df <- dist_pv_df %>%
  filter(year == '2018' | year == '2020')

polls_cvap_vp_df <- merge(polls_cvap_df, dist_pv_df, by = c('st_cd_fips', 'year'))
#table(polls_cvap_vp_df$st_cd_fips)

polls_cvap_vp_df = polls_cvap_vp_df %>%
  select(st_cd_fips, year, state, district_num, dem_cand, rep_cand, DEM, REP, winner_party, RepVotesMajorPercent, DemVotesMajorPercent, RepVotes, DemVotes, cvap) %>%
  mutate(DEM = as.numeric(DEM),
         REP = as.numeric(REP)) %>%
  filter_at(vars(DEM, REP), all_vars(!is.na(.)))
```

From the plot below, we can see that congressional districts, even in the same state, have very different levels of turnout. For example, in Texas, we can see that rural regions have lower turnout (corresponding to purple and blue shades). Urban areas have higher turnout, indicated by brighter shades of red. At first glance, this map seems to confirm an assumption both the Democrats and Republicans operate under: higher turnout helps the Democrats. This is because in rural states that are more conservative, turnout appears to be lower than urban areas. However, this concept may be misleading. For example, even though a rural, conservative district may have low turnout, there may not necessarily be liberal people who do not vote. Instead, it is possible that the district is politically homogenous, and higher turnout would lead to more conservatives voting. We can see this from the plot of contested races in 2018 compared to turnout. There is only a small linear relationship between turnout and Democratic vote share, meaning increased turnout only has a small positive relationship with Democratic success.

```{r, echo=FALSE,warning=FALSE,echo=FALSE, message=FALSE, fig.width=12, fig.height=6, fig.align='center'}
#turnout visualization for 2018

h <- read_csv("/Users/hwalker/Desktop/Senior Fall/Election Analytics/electionanalytics/data/week6/incumb_dist_1948-2020.csv")
# read in cvap

h <- h %>%
  filter(year == 2012 | year == 2014 | year == 2016 | year == 2018 | year == 2020)

# merge
temp <- h %>%
  inner_join(cvap_district, by = c('st_cd_fips', 'year'))

# mutate turnout
temp2 = temp %>%
  mutate(totalvotes = RepVotes + DemVotes,
         turnout = totalvotes/cvap) %>%
  select(state.x,year, district_num, turnout, DemVotesMajorPercent, RepVotesMajorPercent) %>%
  mutate(district_num = as.character(as.numeric(temp$district_num))) %>%
  rename(STATENAME = "state.x",
         DISTRICT = "district_num") %>%
  filter(year =="2018")

get_congress_map = function() {
  tmp_file = tempfile()
  tmp_dir = tempdir()
  download.file("https://cdmaps.polisci.ucla.edu/shp/districts114.zip", tmp_file)
  unzip(zipfile=tmp_file, exdir=tmp_dir)
  fpath = paste(tmp_dir, "districtShapes/districts114.shp", sep="/")
  st_read(fpath)
}
districts = get_congress_map()
districts_turnout <- districts %>% left_join(temp2, by=c("DISTRICT", "STATENAME"))

districts_turnout <- rmapshaper::ms_simplify(districts_turnout, keep = 0.01)

districts_turnout = districts_turnout %>%
  filter(STATENAME != "Alaska" & STATENAME != "Hawaii")
ggplot() + 
  geom_sf(data=districts_turnout,aes(fill=turnout),
          inherit.aes=FALSE,alpha=1) + 
  scale_fill_gradient(low = "blue", high = "red", limits=c(0,1)) +
  coord_sf(xlim=c(-120, -60), ylim=c(25, 50), expand = FALSE) +  
  labs(title = "Turnout by U.S. Congressional District, 2018 Midterm Elections")+
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

```{r, echo=FALSE,warning=FALSE,echo=FALSE, message=FALSE, fig.width=12, fig.height=6, fig.align='center'}
#visualizing 
ggplot(temp2 %>%
  filter(DemVotesMajorPercent < 100 & DemVotesMajorPercent > 0), aes(x = turnout, y = DemVotesMajorPercent)) + 
  geom_point() +
  stat_smooth(method = "lm")
```




```{r,  cache=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
train_data = polls_cvap_vp_df %>% 
  filter(year == '2018' | year == '2020') %>%
  group_by(st_cd_fips, state) %>% 
  filter(n() > 1) %>% # Filtering out single data rows
  group_nest() %>% 
  mutate(data = map(data, ~unnest(., cols = c())))


models = train_data %>% 
  mutate(model_dem = map(data, ~glm(cbind(DemVotes, cvap-DemVotes) ~ DEM, data = .x, family="binomial"))) %>% 
  mutate(model_rep = map(data, ~glm(cbind(RepVotes, cvap-RepVotes) ~ REP, data = .x, family="binomial"))) %>% 
  select(-data)

model_results <- models %>% 
  mutate(standard_error_dem = map_dbl(model_dem, ~summary(.x)$coefficients[, 2][2]),
         standard_error_rep = map_dbl(model_rep, ~summary(.x)$coefficients[, 2][2])
         )
```

*Building a Model*
While turnout may not be useful in a linear model, we can use recent polls as a predictor and then simulate turnout for each party. In this method, we make a generalized linear model for each district, with every recent poll as an observation. Doing 10,000 simulations for each district, using the poll results as the probability of someone voting for a Democrat or Republican, we can get an average margin of victory in each district. From the histograms below, we can see that there are a wide range of outcomes for each seat that we have data for (32 districts in total). Unfortunately, when observing these results, we can see that this model may be problematic. For example, in many instances, neither party gets close to 50% of the vote. Obviously, this does not make sense because in most cases, there is only a Democrat and Republican candidate. While we will eventually predict all the house seats, in our district-level map, it probably makes sense to ignore these results for now. While simulation may be a good idea as we approach our final prediction, turnout is clearly not a useful predictor. 

```{r, echo=FALSE,warning=FALSE,echo=FALSE, message=FALSE, fig.width=12, fig.height=6, fig.align='center'}
test_data <- polls_df %>%
  filter(year == "2022") %>% 
  select(year, state, st_cd_fips, cd_fips, DEM, REP) %>%
  mutate(DEM = as.numeric(DEM),
         REP = as.numeric(REP)) %>%
  filter_at(vars(DEM, REP), all_vars(!is.na(.))) %>% distinct(st_cd_fips, .keep_all = TRUE) %>%
  group_by(st_cd_fips, state) %>% 
  group_nest() %>% 
  mutate(data = map(data, ~unnest(., cols = c())))

pred_2022 <- test_data %>%
  # inner join as there may not be historical models for some districts
  inner_join(models, by = c("state", "st_cd_fips")) %>%
  mutate(pred_dem = map_dbl(.x = model_dem, .y = data, ~predict(object = .x, newdata = as.data.frame(.y),type="response")[[1]])) %>%
  mutate(pred_rep = map_dbl(.x = model_rep, .y = data, ~predict(object = .x, newdata = as.data.frame(.y),type="response")[[1]])) %>%
  select(state, st_cd_fips, pred_dem, pred_rep)

for (x in 1:length(pred_2022$st_cd_fips)){
  sim_D_temp = rbinom(n = 10000, size = 1000, prob = as.numeric(pred_2022[x,3]))
  sim_R_temp = rbinom(n = 10000, size = 1000, prob = as.numeric(pred_2022[x,4]))
  pred_2022$mean_democrat_win_margin[x] = (((sim_D_temp-sim_R_temp)/(sim_D_temp+sim_R_temp))*100)
  t_string = paste("Distribution of results for", pred_2022$state[x],substr(pred_2022$st_cd_fips[x], 3, 4))
  hist(((sim_D_temp-sim_R_temp)/(sim_D_temp+sim_R_temp))*100, main = t_string)
}

library(knitr)

kable(pred_2022)
```




```{r, echo=FALSE,warning=FALSE,echo=FALSE, message=FALSE, fig.width=12, fig.height=6, fig.align='center'}

cols_needed_from_pred2022 = left_join(pred_2022, polls_cvap_vp_df, by="st_cd_fips") %>%
  select(st_cd_fips,state.x, district_num, mean_democrat_win_margin) 

cols_needed_from_pred2022 = cols_needed_from_pred2022 %>%
  mutate(district_num = as.character(as.numeric(cols_needed_from_pred2022$district_num))) %>% unique() %>%
  rename(STATENAME="state.x",
         DISTRICT = "district_num")


districts <- districts %>% left_join(cols_needed_from_pred2022, by=c("DISTRICT", "STATENAME"))

districts_simp <- rmapshaper::ms_simplify(districts, keep = 0.01)

districts_simp = districts_simp %>%
  filter(STATENAME != "Alaska" & STATENAME != "Hawaii")
ggplot() + 
  geom_sf(data=districts_simp,aes(fill=mean_democrat_win_margin),
          inherit.aes=FALSE,alpha=1) + 
  scale_fill_gradient(low = "blue", high = "red", limits=c(-50,50)) +
  coord_sf(xlim=c(-120, -60), ylim=c(25, 50), expand = FALSE) +  
  labs(title = "Simulated Win Margin by Available U.S. Congressional District, 2018")+
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```


*Looking Ahead*

Going forward, I think pooling all the past congressional races (for which we have data) makes the most sense for predicting the upcoming midterms. While grouping the models by district makes sense in theory, I have not added any data that helps describe the characteristics of a district. For example, we have yet to incorporate levels of education, or how rural or urban the district is. Instead, like in this exercise, I made a model for each district, but each one had very few observations, leading to results that did not make sense. 

